---
title: "class 11Structural Bioinformatics"
format: html
editor: visual
---



>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have? 

```{r}
expr <- read.table("textfile.txt")

head (expr)


```

```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```



>Q14 Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.


lets make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
  
```
I could infer that A|A has a higher expression than G|G. The reduced expression of G|G could show that
SNP does effect the expression of ORMLDL3 since  it altered the gene expression level, potentially influencing the disease function and immune response. 
